Convergence rate of MAP estimates \\ for the exponential family

### 2021-04-14 main.tex
New lower bound on the logarithm

\begin{example}
	For $\cN(0,\sigma^2)$, both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_n^2 ; \sigma_*^2) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}
In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) \\
	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
as illustrated in Figure~\ref{fig:phi}.
\end{example}

@@

\begin{example}
	\paragraph{Exact computation}
	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \sum_i X_i^2 /n$. The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$. Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation 
	\begin{align}
		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
		= \expect{\frac{n}{\chi^2(n)} - 1} =
		\begin{cases}
			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
		\end{cases}
	\end{align}
	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
	\begin{align}
		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
	\end{align}
	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
	Consequently the suboptimality of the MLE has a closed form solution
	\begin{align}
	\expect{\bregmanconj( \mu_*; \mu_n) }
	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
	& =
	\begin{cases}
		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
	\end{cases}
	\end{align}
	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
	\begin{align}
		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
	\end{align}
	  For $n\geq 3$,
	\begin{align}
		\inv{n-2} - \inv{n}
		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
		\leq \inv{n-2} - \inv{2 n} \\
		\iff
			\frac{2}{n(n-2)}
			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
			\leq \inv{2n} +\frac{2}{n(n-2)}
			\label{eq:tight_bounds}
	\end{align}
	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
	\begin{align}
		\expect{\bregmanconj( \mu_*; \mu_n) }
		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
	\end{align}
	Unfortunately, formulas for the MAP seem to be much less nice. There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. Again the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.
\end{example}

@@

\begin{example}
		We need a formula  for $\E[\MAPt]$ the expectation of the natural MAP estimate
	\begin{align}
		\E[\MAPt] &= -\frac{n+n_0}{2}\expect{\inv{n_0 \mu_0 + \sum_i X_i^2}}  \\
		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{a + \chi^2(n)}}
	\end{align}
	where $\sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
	\begin{align}
			a=n_0 \frac{\mu_0}{\mu^*} \; .
	\end{align}
	This expectations is not simple to solve, but using wolfram, we get  the formula :
	\begin{align}
		\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])
	\end{align}
	where the \href{https://dlmf.nist.gov/8.19}{generalized exponential integral function} is defined as 
	\begin{align}
		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
	\end{align}
	Now our goal is to bound this function with simpler functions.
	Fortunately, mathematicians have been working on these integrals for decades.
	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8#E1}{verifies the bound}
	\begin{align}
		\half \log(1 + \frac{2}{x}) 
		\leq e^x E_1(x) 
		\leq \log(1+\inv{x}) \leq \inv{x} \; .
	\end{align}
	For larger $n$, we have \href{https://dlmf.nist.gov/8.19#E21}{the general bound}
	\begin{align}
		\inv{x+n}
		\leq e^x E_k(x) 
		\leq \inv{x + k - 1}
	\end{align}
	so that 
	\begin{align}
		\inv{a+n}
		\leq \expect{\inv{a + \chi^2(n)}}
		\leq \inv{a + n - 2}
	\end{align}
	so that
	\begin{align}
		\frac{n +n_0}{2 \mu^* (a + n)}	
		&\leq - \E[\MAPt]
		\leq \frac{n +n_0}{2 \mu^* (a + n - 2)}	\\
		\iff \inv{2 \mu_n}
		&\leq - \E[\MAPt]
		\leq \inv{2 \mu_n - \frac{4 \mu^*}{n + n_0}}
		\label{eq:theta_expectation_bound}
	\end{align}
	because $\frac{\mu^* (a + n)}{n +n_0} = \mu_n$. Remark how we almost bounded the deviation between $\E[\MAPt]$ and $\natp_n$.
\end{example}

	\begin{figure}[ht]
    	\centering
	    \includegraphics[width=.7\textwidth]{figs/wolfram.png}
    	\label{fig:wolfram}
	\end{figure}

@@

\begin{example}
	\begin{align}
		\bregmanconj(\mu^*; \mu_n)
		 &= \frac{\mu^*}{\mu_n} -1  - \log \frac{\mu^*}{\mu_n}  \\
		 & = \frac{n+n_0}{a + n} - 1 - \log \frac{n+n_0}{a +n} \\
		 & = \frac{n_0 - a}{a + n} - \log \frac{n+n_0}{a +n} \\		 
	\end{align}
\end{example}

@@

>>>>

\begin{example}
	For $\cN(0,\sigma^2)$, both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_n^2 ; \sigma_*^2) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}
In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) \\
	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
\begin{align}
	1 - \inv{z} \leq \log(z) 
	\label{eq:log_bound}\\
	\implies \phi(z) &\leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z}
\end{align}
as illustrated in Figure~\ref{fig:phi}.
\end{example}

@@

\begin{example}
	\paragraph{MLE exact computation}
	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \sum_i X_i^2 /n$. The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$. Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation 
	\begin{align}
		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
		= \expect{\frac{n}{\chi^2(n)} - 1} =
		\begin{cases}
			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
		\end{cases}
	\end{align}
	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
	\begin{align}
		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
	\end{align}
	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
	Consequently the suboptimality of the MLE has a closed form solution
	\begin{align}
	\expect{\bregmanconj( \mu_*; \mu_n) }
	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
	& =
	\begin{cases}
		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
	\end{cases}
	\end{align}
	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
	\begin{align}
		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
	\end{align}
	  For $n\geq 3$,
	\begin{align}
		\inv{n-2} - \inv{n}
		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
		\leq \inv{n-2} - \inv{2 n} \\
		\iff
			\frac{2}{n(n-2)}
			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
			\leq \inv{2n} +\frac{2}{n(n-2)}
			\label{eq:tight_bounds}
	\end{align}
	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
	\begin{align}
		\expect{\bregmanconj( \mu_*; \mu_n) }
		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
	\end{align}
	Unfortunately, formulas for the MAP seem to be much less nice. There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. Again the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.
\end{example}

\begin{example}
	\paragraph{MAP upper bound.}
	Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ but it does for $\expect{\frac{\mu^*}{\MAPm}} = \expect{\frac{\MAPt}{\natp^*}}$. 
	To start, let us plug in the definition of $\MAPm$
	\begin{align}
		\E[\MAPt] &= -\frac{n+n_0}{2}\expect{\inv{n_0 \mu_0 + \sum_i X_i^2}}  \\
		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{a + \chi^2(n)}}
	\end{align}
	where $\sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
	\begin{align}
			a=n_0 \frac{\mu_0}{\mu^*} \; .
	\end{align}
	This expectations is not simple to solve, but using wolfram, we get  the formula :
	\begin{align}
		\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])
		\label{eq:wolfram_input}
	\end{align}
	where the \href{https://dlmf.nist.gov/8.19}{generalized exponential integral function} is defined as 
	\begin{align}
		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
	\end{align}
	Now our goal is to bound this function with simpler functions.
	Fortunately, mathematicians have been working on these integrals for decades.
	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8#E1}{verifies the bound}
	\begin{align}
		\half \log(1 + \frac{2}{x}) 
		\leq e^x E_1(x) 
		\leq \log(1+\inv{x}) \leq \inv{x} \; .
	\end{align}
	For larger $n$, we have \href{https://dlmf.nist.gov/8.19#E21}{the general bound}
	\begin{align}
		\inv{x+n}
		\leq e^x E_k(x) 
		\leq \inv{x + k - 1}
	\end{align}
	so that 
	\begin{align}
		\inv{a+n}
		\leq \expect{\inv{a + \chi^2(n)}}
		\leq \inv{a + n - 2}
	\end{align}
	so that
	\begin{align}
		\frac{n +n_0}{2 \mu^* (a + n)}	
		&\leq - \E[\MAPt]
		\leq \frac{n +n_0}{2 \mu^* (a + n - 2)}	\\
		\iff \inv{2 \mu_n}
		&\leq - \E[\MAPt]
		\leq \inv{2 \mu_n - \frac{4 \mu^*}{n + n_0}}	\\
		\iff \frac{\mu^*}{\mu_n}
		&\leq \expect{\frac{\mu^*}{\MAPm}} 
		\leq \frac{\mu^*}{\mu_n - \frac{4 \mu^*}{n + n_0}}  
		\label{eq:theta_expectation_bound}
	\end{align}
	because $\frac{\mu^* (a + n)}{n +n_0} = \mu_n$.
Now using the bound \eqref{eq:log_bound} on the logarithm, we get
\begin{align}
	2\expect{\bregmanconj(\mu^*; \MAPm)} 
	&\leq \expect{\frac{\mu^*}{\MAPm}}  + \expect{\frac{\MAPm}{\mu^*}} - 1 \\
	&\leq \frac{\mu^*}{\mu_n - \frac{4 \mu^*}{n + n_0}}   + \frac{\mu_n}{\mu^*} - 1 \\
	&= \frac{n_0 + n}{a +n - 2} + \frac{a+n}{n_0 + n} - 2 \\
	&= \frac{n_0 - a + 2}{a +n - 2} + \frac{a - n_0}{n_0 + n} \\
	&= \frac{2}{a+n-2} + \frac{(n_0 - a)^2}{(a+n-2)(n_0 + n)}
\end{align}
where we remind you that $a=n_0\frac{\mu_0}{\mu^*}$. This inequality highlights a clear variance-bias decomposition, to compare with the MLE.
\end{example}

	\begin{figure}[ht]
    	\centering
	    \includegraphics[width=.7\textwidth]{figs/wolfram.png}
	    \caption{Output of Wolfram when queried for the integral \eqref{eq:wolfram_input}}
    	\label{fig:wolfram}
	\end{figure}

@@

@@

\begin{example}
We can use the classical lower bound on the logarithm to upper bound the pure bias term 
	\begin{align}
		\bregmanconj(\mu^*; \mu_n)
		 &= \frac{\mu^*}{\mu_n} -1  - \log \frac{\mu^*}{\mu_n}  \\
		 &\leq \frac{(\mu^*- \mu_n)^2}{2 \mu^* \mu_n} 
		 = \half ( \frac{\mu^*}{\mu_n} + \frac{\mu_n}{\mu^*	}) - 1 \\		 
		 & =  \half (\frac{n_0 + n}{a + n} -1 + \frac{a +n}{n_0+n} -1) \\
		 & =  \half (\frac{n_0 - a}{a+n} + \frac{a - n_0}{n_0 + n}) \\
		 & = \half[n_0-a] (\inv{a+n} - \inv{n_0+n}) \\
		 &= \half[n_0-a] \frac{n_0 + n - a- n}{(a+n)(n_0+n)} \\
		 &= \frac{(n_0-a)^2}{2(a+n)(n_0+n)}
	\end{align}
	which is a $O(n^{-2})$ term, worth $0$ when $n_0=a$, eg $\mu_0=\mu^*$, exactly as we expected.
\end{example}

@@

\subsection{Putting it all together}
\begin{example}
\begin{align}
	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	\leq \bias{\frac{(n_0-a)^2}{2(a+n)(n_0+n)}}
	+ \frac{\max(n_0 - a , 0)}{(a+n)(a+n-2)}
	+ \variance{\frac{1}{a+n-2}}
\end{align}
\end{example}
